Dynamic parameter estimation in the nested multinomial logit choice model

### 2021-08-08 paper/paper.tex
rev

\begin{abstract}
This paper considers the dynamic estimation of the nested multinomial logit choice model under a tight computational constraint. I provide an approximation scheme for the model parameters that updates the current estimate according to an intuitive step rule. Stochastic convergence is guaranteed under the Robbins--Monro condition. Numerical simulations demonstrate the technique's feasibility.
\end{abstract}

@@

%%Research highlights
\begin{highlights}
\item This paper presents a technique for estimating the parameters of the nested multinomial logit choice model in a dynamic computational setting. 
\item The technique is based on a legible step rule whose convergence is guaranteed by classical results in stochastic approximation. 
\end{highlights}

@@

\section{Introduction}
In a recent paper, \cite{honguyen2021} articulate the need for \emph{dynamic} choice model fitting techniques that can update the current estimate in light of new observations of consumer behavior without solving a full optimization problem over the perturbed dataset. That paper concerned nonparametric choice models, which are hard to estimate even in a static setting \cite[][]{rusmevichientong2006, farias2013}. On the other hand, while \emph{parametric} choice models cannot accommodate every distribution of consumer preference orders \cite[][]{keane1997}, they offer a substantial advantage in the computational tractability of both model fitting and assortment optimization tasks \cite[][]{bunch1987, davis2014}. To establish parity in the larger debate over the relative merits of parametric and nonparametric choice models, it is worth examining whether this tractability advantage is equally pronounced in the dynamic setting.

The present paper answers in the affirmative by providing a dynamic approximation scheme for the multinomial logit choice model and its nested variant in a tightly constrained computational environment called the data stream. When a new observation of a consumer choice arrives, the technique proposed here updates the current parameter estimate according to a simple rule with a sequence of decreasing step sizes. Stochastic convergence is guaranteed under the Robbins--Monro condition, and can be improved by tuning step parameters and utilizing a rolling average. I conclude by demonstrating the feasibility of the technique through numerical simulations.

@@

In the multinomial logit (MNL) choice model, each consumer $t$'s preference order over the products in $\mathcal{A}$ is determined by the sum of a systematic product preferability parameter $\delta_i$ and a noise term $\epsilon_{it}$. The noise terms have an iid extreme value distribution, and the consumer chooses the product in $\mathcal{A}$ for which $\delta_i + \epsilon_{it}$ is greatest. It can be shown that the probability of a given consumer choosing product $i$ from the assortment $\mathcal{A}$ is
\begin{equation}\label{mnlchoiceprobability}
\operatorname{Pr}\left[\,\text{choose } i \;|\;\text{assortment }\mathcal{A}\,\right] =\frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j}
\end{equation}
The value of this equation does not change when each $\delta_i$ is perturbed by the same constant; hence, we may assume without loss of generality that $\sum_{i=1}^n \delta_i = 0$.

@@

Now consider the NMNL choice model. Each consumer $t$ chooses a product from an assortment $\mathcal{A}_t$, and we may write the consumer's choice as $(x_t, y_t)$, where $y_t$ is the product chosen and $x_t$ is the nest to which $y_t$ belongs. (This requires that the nests be allocated in advance.) Again, $\mathcal{A}_t$ as a list of (nest, product) tuples $(k, i)$. Let $\mathcal{A}_t^{\text{nests}}$ denote the set of nests represented in the assortment $\mathcal{A}_t$, let $\mathcal{A}_{t}^{\text{products}}$ denote the set of products that are in both nest $x_t$ and the assortment $\mathcal{A}_t$, and abbreviate the parameters $\sigma_{x_t}$ and $\delta_{y_t}$ to $\sigma_t$ and $\delta_t$. Then the likelihood function is
\begin{equation}L(\sigma, \delta) = \prod_{t=1}^T \frac{\exp \sigma_t}{\sum_{l\in \mathcal{A}_t^{\text{nests}}} \exp \sigma_l} \cdot \frac{\exp \delta_t}{\sum_{k \in \mathcal{A}_{t}^{\text{products}}} \exp \delta_k}\end{equation}
and is concave in the logarithm.

@@

\section{Parameter estimation in the data stream framework}
The dynamic parameter estimation scheme considered in this paper applies to an online setting called the data stream. In the data stream, we are only allowed to store our \emph{current} parameter estimate $\delta$ and a state variable $t$ representing the number of observations used to form the current estimate. When a new observation $(\mathcal{A}_t, y_t)$ comes in, we can update the stored variables, but we cannot store the observation itself. %A simple example of a parameter estimation scheme in the data stream framework is the updating sample mean $\bar x^{(t+1)} = \frac{1}{t+1} \left( x_{t+1} + t \bar x^{(t)}\right)$. This calculation eliminates the need to store past observations of the random variable $X$, and each update is $\mathcal{O}(1)$ instead of the $\mathcal{O}(t)$ operations required to recalculate the mean over the full dataset. The MNL parameter update described below offers an analogous computational advantage.

@@

\subsection{Why the parameter update converges}
Consider the multivariate root-finding problem \[\text{find } \theta^*: F(\theta^*) = 0\]
and the iterative scheme \[\theta^{(t+1)} = \theta^{(t)} + \frac{\alpha}{t^r} F\left(\theta^{(t)} \right)\]
where $\alpha >0$ and $r \in (0, 1]$ are step parameters. When $F: \mathbb{R}^n \mapsto \mathbb{R}^n$ is Lipschitz continuous and its Jacobian is negative definite, such an iterative scheme is known to converge to the root, if it exists (\citealp[][Theorem 11.6]{nocedal2006}; \citealp[][Proposition 1.2.4]{bertsekas1999}). A typical example occurs when using gradient ascent to maximize a strictly concave function whose gradient is $F$.

@@

Let $\delta$ denote the current parameter estimate and let $\delta^*$ denote the true, unknown parameters of the underlying choice model. When the probability of offering each consideration set $\mathcal{A}$ is $\operatorname{Pr[\mathcal{A}]} > 0$, then the expected value of $Z$ has
\begin{align}
\mathbb{E}\left[Z(\delta)_i \right] &= \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \mathbb{E}\left[Z(\delta)_i~|~\mathcal{A}\right] \\
&=   \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \left(\operatorname{Pr}\left[\,\text{choose } i\;|\;\mathcal{A}\,\right] - \frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j} \right) \\
&= \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\left(\frac{\exp  \delta_i^*}{\sum_{j\in \mathcal{A}} \exp \delta_j^*} - \frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j} \right) \\
&= 0
\end{align}
if and only if $\delta^* = \delta$. Hence, the counterpart to $F(\theta)$ is the expected value of the random function $Z(\delta)$, while the counterpart to the experiment $\hat F(\theta)$ is the realization $Z^{(t)}(\delta)$. Next, observe that the Jacobian of $\mathbb{E}\left[Z(\delta) \right]$ is negative definite: The first term of the summand is summand is constant, and the second is identical in form to the gradient of the logarithm of the likelihood function given in Equation \eqref{mnllikelihood}, which is concave. Thus, we conclude that the iterative procedure is a Robbins--Monro algorithm and converges in expectation to $\delta^*$.

@@

\begin{figure}
\begin{center}\includegraphics[width=\textwidth]{../plots/param-est-dynamic-nonnested.png}\end{center}
\captionsetup{singlelinecheck=off}
    \caption[.]{Dynamic estimation of MNL choice parameters for a collection of $n = 7$ products. True preferability parameters $\delta$ and the initial estimate $\delta^{(1)}$ were drawn independently from a normal distribution and centered at zero. At times $t = 1 \dots 25000$, a random assortment consisting of between two and five products was drawn, a consumer chose a product using the MNL choice model, and the parameter estimates were updated according to the MNL parameter update (Equation \ref{mnlparameterupdate}). The parameter estimate at each iteration is shown in a light stroke, while the darker lines show a moving average over the previous 500 estimates. The step parameters are $\alpha = 0.01$ and $r = 0.05$.}
\label{param-est-dynamic-nonnested}
\end{figure}

\begin{figure}
\begin{center}\includegraphics[width=\linewidth, ]{../plots/param-est-dynamic-nested.png}\end{center}
\captionsetup{singlelinecheck=off}
    \caption[.]{Dynamic estimation of MNL choice parameters for a collection of $mn = 12$ products divided across $m=4$ nests. True preferability parameters $\sigma$ and $\delta$ were drawn independently from a normal distribution and centered at zero, while the initial parameter estimates were set to zero. At times $t = 1 \dots 60000$, a random assortment consisting of between two and five products was drawn, a consumer chose a product using the MNL choice model, and the parameter estimates were updated according to the NMNL parameter update (Equation \ref{nmnlparameterupdate}). The parameter estimate at each iteration is shown in a light stroke, while the darker lines show a moving average over the previous 1200 estimates. The step parameters are $\beta = 0.01$, $\alpha = 0.03$, and $r = 0.05$. The first pane shows the time series of $\sigma$ estimates, while the second shows the entries of $\delta$, color-coded by nest membership.}
\label{param-est-dynamic-nested}
\end{figure}

\begin{figure}
\begin{center}\includegraphics[width=\linewidth, ]{../plots/param-est-dynamic-nested-as-static.png}\end{center}
\captionsetup{singlelinecheck=off}
    \caption[.]{Performance of the dynamic parameter estimate versus maximum likelihood estimation. Preferability parameters for $mn = 120$ products divided into $m=10$ nests were drawn from the normal distribution and centered at zero. $T=50000$ assortments and consumer choices were generated according to the NMNL model. In the upper two panes, the observations were treated as a time series, and I report the average of the final 1000 estimates produced by the dynamic parameter update with step parameters $\beta = 0.01$, $\alpha = 0.03$, and $r = 0.05$. In the lower two panes, maximum likelihood estimation was used to estimate the parameters. Dynamic parameter estimation offers acceptable accuracy at substantially lower computational cost.}
\label{param-est-dynamic-nested-as-static}
\end{figure}

>>>>

\begin{abstract}
This paper considers the dynamic parameter estimation of the nested multinomial logit choice model under a tight computational constraint. I provide an approximation scheme for the model parameters that updates the current estimate according to an intuitive step rule. Stochastic convergence is guaranteed under the Robbins--Monro condition. Numerical simulations demonstrate the technique's feasibility.
\end{abstract}

@@

%%Research highlights
%\begin{highlights}
%\item This paper presents a technique for estimating the parameters of the nested multinomial logit choice model in a dynamic computational setting with a tight data limit. 
%\item The technique is based on a legible step rule whose convergence is guaranteed by classical results in stochastic approximation.
%\item Numerical simulations demonstrate the technique's feasibility.
%\end{highlights}

@@

\section{Introduction}
In a recent paper, \cite{honguyen2021} articulate the need for \emph{dynamic} choice modeling that can incorporate new observations of consumer behavior without solving a full optimization problem over the perturbed dataset. That paper concerned nonparametric choice models, which are hard to estimate even in a static setting \cite[][]{rusmevichientong2006, farias2013}. On the other hand, while \emph{parametric} choice models cannot accommodate every distribution of consumer preference orders \cite[][]{keane1997}, they offer a substantial advantage in the computational tractability of both model fitting and assortment optimization tasks \cite[][]{bunch1987, davis2014}. To establish parity in the  debate over the relative merits of parametric and nonparametric choice models, it is worth examining whether this tractability advantage is equally pronounced in the dynamic setting.

The present paper answers in the affirmative by providing a dynamic approximation scheme for the multinomial logit choice model and its nested variant in a tightly constrained computational environment called the data stream. When a new observation of a consumer choice arrives, the technique proposed here updates the current parameter estimate according to an intuitive rule with a sequence of decreasing step sizes. I show that stochastic convergence is guaranteed under the Robbins--Monro condition and the convergence rate can be honed by tuning step parameters and utilizing a rolling average. I conclude by demonstrating the feasibility of the technique through numerical simulations.

@@

In the multinomial logit (MNL) choice model, each consumer $t$'s preference order over the products in $\mathcal{N}$ is determined by the sum of a systematic product preferability parameter $\delta_i$ and a noise term $\epsilon_{it}$. The noise terms have an iid extreme value distribution, and the consumer chooses the product in $\mathcal{A}$ for which $\delta_i + \epsilon_{it}$ is greatest. It can be shown that the probability of a given consumer choosing product $i$ from the assortment $\mathcal{A}$ is
\begin{equation}\label{mnlchoiceprobability}
\operatorname{Pr}\left[\,\text{choose } i \;|\;\text{assortment }\mathcal{A}\,\right] =\frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j}
\end{equation}
The value of this equation does not change when each $\delta_i$ is perturbed by the same constant; hence, we may impose the normalization rule $\sum_{i=1}^n \delta_i = 0$ without loss of generality.

@@

Now consider the NMNL choice model. Each consumer $t$ chooses a product from an assortment $\mathcal{A}_t$, and we may write the consumer's choice as $(x_t, y_t)$, where $y_t$ is the product chosen and $x_t$ is the nest to which $y_t$ belongs. Again, $\mathcal{A}_t$ may be written as a list of (nest, product) tuples $(k, i)$. Let $\mathcal{A}_t^{\text{nests}}$ denote the set of nests represented in the assortment $\mathcal{A}_t$, let $\mathcal{A}_{t}^{\text{products}}$ denote the set of products that are in both nest $x_t$ and the assortment $\mathcal{A}_t$, and abbreviate the parameters $\sigma_{x_t}$ and $\delta_{y_t}$ to $\sigma_t$ and $\delta_t$. Then the likelihood function is
\begin{equation}L(\sigma, \delta) = \prod_{t=1}^T \frac{\exp \sigma_t}{\sum_{l\in \mathcal{A}_t^{\text{nests}}} \exp \sigma_l} \cdot \frac{\exp \delta_t}{\sum_{k \in \mathcal{A}_{t}^{\text{products}}} \exp \delta_k}\end{equation}
which is concave in the logarithm.

@@

\section{Parameter estimation in the data stream framework}
The dynamic scheme considered in this paper estimates NMNL parameters in online setting called the data stream. In the data stream, we are only allowed to store our \emph{current} parameter estimate and a state variable $t$ representing the number of observations used to construct the current estimate. When a new observation $(\mathcal{A}_t, y_t)$ comes in, we can update the stored variables, but we cannot store the observation itself. %A simple example of a parameter estimation scheme in the data stream framework is the updating sample mean $\bar x^{(t+1)} = \frac{1}{t+1} \left( x_{t+1} + t \bar x^{(t)}\right)$. This calculation eliminates the need to store past observations of the random variable $X$, and each update is $\mathcal{O}(1)$ instead of the $\mathcal{O}(t)$ operations required to recalculate the mean over the full dataset. The MNL parameter update described below offers an analogous computational advantage.

@@

\subsection{Why the parameter update converges}
Consider the multivariate root-finding problem \[\text{find } \theta^*: F(\theta^*) = 0\]
and the iterative scheme \[\theta^{(t+1)} = \theta^{(t)} + \frac{\alpha}{t^r} F\left(\theta^{(t)} \right)\]
where $\alpha >0$ and $r \in (0, 1]$ are step parameters. When $F: \mathbb{R}^n \mapsto \mathbb{R}^n$ is Lipschitz continuous and its Jacobian is negative definite, such an iterative scheme is known to converge to the root, if it exists (\citealp[][thm. 11.6]{nocedal2006}; \citealp[][prop. 1.2.4]{bertsekas1999}). A typical example occurs when using gradient ascent to maximize a strictly concave function whose gradient is $F$.

@@

Let $\delta$ denote the current parameter estimate and let $\delta^*$ denote the true, unknown parameters of the underlying MNL choice model. When the probability of offering each consideration set $\mathcal{A}$ is $\operatorname{Pr[\mathcal{A}]} > 0$, then the expected value of $Z$ has
\begin{align}
\mathbb{E}\left[Z(\delta)_i \right] &= \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \mathbb{E}\left[Z(\delta)_i~|~\mathcal{A}\right] \\
&=   \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\operatorname{Pr}\left[\mathcal{A}\right] \cdot \left(\operatorname{Pr}\left[\,\text{choose } i\;|\;\mathcal{A}\,\right] - \frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j} \right) \\
&= \sum_{\substack{\mathcal{A} \in 2^{\mathcal{N}}: \\ i \in \mathcal{A}}}\left(\frac{\exp  \delta_i^*}{\sum_{j\in \mathcal{A}} \exp \delta_j^*} - \frac{\exp \delta_i}{\sum_{j\in \mathcal{A}} \exp \delta_j} \right) \\
&= 0
\end{align}
if and only if $\delta^* = \delta$. Hence, the counterpart to $F(\theta)$ is the expected value of the random function $Z(\delta)$, while the counterpart to the experiment $\hat F(\theta)$ is the realization $Z^{(t)}(\delta)$. Next, observe that the Jacobian of $\mathbb{E}\left[Z(\delta) \right]$ is negative definite: The first term of the summand is constant, and the second is identical in form to the gradient of the logarithm of the likelihood function given in Equation \eqref{mnllikelihood}, which is concave. Thus, we conclude that the iterative procedure is a Robbins--Monro algorithm and converges in expectation to $\delta^*$.

@@

\begin{figure}
\begin{center}\includegraphics[width=\textwidth]{../plots/param-est-dynamic-nonnested.png}\end{center}
\captionsetup{singlelinecheck=off}
    \caption[.]{(Color.) Dynamic estimation of MNL choice parameters for a collection of $n = 7$ products. True preferability parameters $\delta$ and the initial estimate $\delta^{(1)}$ were drawn independently from a normal distribution and centered at zero. At times $t = 1 \dots 25000$, a random assortment consisting of between two and five products was drawn, a consumer chose a product using the MNL choice model, and the parameter estimates were updated according to the MNL parameter update (Equation \ref{mnlparameterupdate}). The parameter estimate at each iteration is shown in a light stroke, while darker lines indicate a moving average over the previous 500 estimates. The step parameters are $\alpha = 0.01$ and $r = 0.05$.}
\label{param-est-dynamic-nonnested}
\end{figure}

\begin{figure}
\begin{center}\includegraphics[width=\linewidth, ]{../plots/param-est-dynamic-nested.png}\end{center}
\captionsetup{singlelinecheck=off}
    \caption[.]{(Color.) Dynamic estimation of MNL choice parameters for a collection of $mn = 12$ products divided across $m=4$ nests. True preferability parameters $\sigma$ and $\delta$ as well as the initial parameter estimates were drawn independently from a normal distribution and centered at zero. At times $t = 1 \dots 60000$, a random assortment consisting of between two and five products was drawn, a consumer chose a product using the MNL choice model, and the parameter estimates were updated according to the NMNL parameter update (Equation \ref{nmnlparameterupdate}). The parameter estimate at each iteration is shown in a light stroke, while darker lines indicate a moving average over the previous 1200 estimates. The step parameters are $\beta = 0.01$, $\alpha = 0.03$, and $r = 0.05$. The first pane shows the time series of $\sigma$ estimates, while the second shows the entries of $\delta$, color-coded by nest membership.}
\label{param-est-dynamic-nested}
\end{figure}

\begin{figure}
\begin{center}\includegraphics[width=\linewidth, ]{../plots/param-est-dynamic-nested-as-static.png}\end{center}
\captionsetup{singlelinecheck=off}
    \caption[.]{(Color.) Performance of the dynamic parameter estimate versus maximum likelihood estimation. Preferability parameters for $mn = 120$ products divided into $m=10$ nests were drawn from the normal distribution and centered at zero. $T=50000$ assortments and consumer choices were generated according to the NMNL model. In the upper two panes, the observations were treated as a time series, and I report the average of the final 1000 estimates produced by the dynamic parameter update with step parameters $\beta = 0.01$, $\alpha = 0.03$, and $r = 0.05$. In the lower two panes, maximum likelihood estimation was used to estimate the parameters. Dynamic parameter estimation offers acceptable accuracy at substantially lower computational cost.}
\label{param-est-dynamic-nested-as-static}
\end{figure}
